---
title: "Data Manipulation"
author: "Kaleb J. Frierson"
date: "2024-09-19"
output: 
  github_document:
    toc: TRUE
---

# Notes

Major steps: 

Select relevant variables, filter out unnecessary obs, create new variables or change existing ones, arrange in an easy to digest format. Make it make sense to you and your research question. Each of these steps has a function within the dplyr package. 

dplyr comes with tidyverse. So just call that library. 

  The idea that you have one function per step is an intentional choice       within dplyr. 
  
  Dataframe centric: start with a dataframe and end with a dataframe. One     goes in and one goes out. That is how it works in tidyverse every time. 
  
  This allows you to pipe together collections of operations that gets you    to the dataset that you need. Piece meal resulting in big thing that you    need. Pipes came from magrittr package; loaded by everything in the         tidyverse. There is an updated one that is built into R!

So, what are pipe operators? 

  Steps to day: 
    
    Wake up, brush teeth, do data science
    
In "R" I can nest these actions: 

happy_kaleb = do_ds(brush_teeth(wake_up(asleep_kaleb)))

This is hard to read! Alternatively you can name a bunch of intermediate objects that you don't need and you end up with a bunch of dataframes and you might not know which to use. 

Piping says start with one thing go to another then another: 

happy_kaleb = 
wake_up(asleep_kaleb) |>
brush_teeth() |>
do_ds()

Read |> as "and then"; you have all of these steps linked together!

Per Jeff: be greatful that you don't have to appreciate how great piping is!

# Setup

```{r library calling, echo = FALSE, message = FALSE}
library(tidyverse)
```

```{r}
litters_df = 
  read_csv("data_import_examples/FAS_litters.csv", na = c("NA","","."))

litters_df = janitor::clean_names(litters_df) 
```

```{r}
pups_df = 
  read_csv("data_import_examples/FAS_pups.csv", na=c("NA","","."))

pups_df = janitor::clean_names(pups_df)
```

# Data Manipulation

## select

Use select to select variables: 

```{r}
select(litters_df, group, litter_number, gd0_weight)
```

```{r}
select(litters_df, group:gd18_weight)
```

```{r}
select(litters_df, -pups_survive)
```

```{r}
select(litters_df, -(group:gd18_weight)) 
```


```{r}
select(litters_df, starts_with("gd")) 

select(litters_df, contains("birth"))
```

```{r}
select(litters_df, GROUP=group)

rename(litters_df, GROUP=group)
```
 using rename above lets you keep the rest of the variables. 
 
```{r}
select(litters_df, litter_number, gd0_weight, everything())

relocate(litters_df, litter_number, gd0_weight)
```

Using relocate you don't have to state everything after rearranging/stating what you want first two columns to be. 

Learning assessment: 

```{r}

```

## filter

Filter works on rows/observations. 

```{r}
filter(litters_df, gd_of_birth == 20) 
```

This is how you can ask whether something is true about a variable! 

```{r}
filter(litters_df, pups_born_alive >8) 
filter(litters_df, pups_born_alive >=8) 
filter(litters_df, pups_born_alive !=9) 
```

What about character variables? 

```{r}
filter(litters_df, group=="Low8")
filter(litters_df, group %in% c("Low7","Low8"))
```
```{r}
filter(litters_df, group %in% c("Low7","Low8"), pups_born_alive ==8)
```
Remember, anything with variables is CASE SENSITIVE!!!

Below: you can drop everything or you can specift what to drop. 

```{r}
drop_na(litters_df)
drop_na(litters_df, gd0_weight)
```
^ lets you filter out rows with missing information

Learning assessment: 

```{r}
filter(pups_df, sex == 1) 
```

## mutate



## arrange
## piping 











